---
title: "Example Analysis"
output: html_notebook
author: Anna Bird
date: 10.04.21
---



### Purpose of the script: demonstrate coding skills using *in silico*-generated (non-proprietary) flow cytometry data

<br><br>

#### How to use this script
1) run the script Ctrl+Alt+R
2) Click "Preview" button above to see HTML Analysis Report

##### Script Inputs 
- "example FACS data.xlsx" This includes raw data & metadata for the FACS experiment

##### Script Outputs
- Report output from this .Rmd ________________Primary Analysis (cf "How to use this script")
- "antigen responses to drug treatments.pdf" __________ Supplementary exported plots
- "120901 FACSstats CleanedMerged.xlsx" __________ exported tidied data set


<br><br>

### Assessing the efficacy of several compounds to improve the maturation & function of Cell Type X


<br><br>

```{r adjust settings & load libraries, include = FALSE}

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 120),
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  echo = FALSE,
  collapse = TRUE
)

library(ape)
library(colormap)
library(data.table)
library(datasets)
library(devtools)
library(extrafont)
library(extrafontdb)
library(ggforce)
library(ggplot2)
require(ggpointdensity)
library(ggrepel)
library(grid)
library(gridExtra)
library(gtable)
library(hms)
library(knitr)
library(lattice)
library(lubridate)
library(Matrix)
library(memisc)
library(MASS)
library(patchwork)
library(plyr)
library(RColorBrewer)
library(readxl)
library(recipes)
library(remotes)
library(rlang)
library(scales)
library(stringr)
library(stringi)
library(tidyr)
library(tidyselect)
library(tidyverse)
library(tidymodels)
library(viridis)
library(writexl)
library(vroom)
library(glue)
library(tidymodels)
library(viridis)

extrafont::loadfonts(device="pdf")
extrafont::loadfonts(device="postscript")
theme_set(theme_bw(base_family = "Helvetica"))
theme_update(strip.background = element_rect(color="white", fill="#FFFFFF", size=0), 
          strip.text = element_text(colour = 'black', size = 12, face = "bold", hjust = 0),
          plot.tag = element_text(color = "ivory", hjust = 1, size = 8))

allDuplicated <- function(vec){
  front <- duplicated(vec)
  back <- duplicated(vec, fromLast = TRUE)
  all_dup <- front + back > 0
  return(all_dup)
}

toGreek <- function(x) {    #use library(stringi) ... 

   x <- gsub("alpha", "\u03b1", x)              # English to unicode
   x <- gsub("beta", "\u03b2", x)
   x <- gsub("gamma", "\u03b3", x)
   #x <- gsub("delta", "\u03b4", x)
   x <- gsub("epsilon", "\u03b5", x)
   x <- gsub("zeta", "\u03b6", x)
   x <- gsub("ug/mL", "\u03bcg/mL", x) 
   x <- gsub("uL", "\u03bcL", x)

}

toGreekLiteral <- function(x) {   

   x <- gsub("ug/mL", "μg/mL", x) 
   x <- gsub("alpha", "α", x)            
   x <- gsub("beta", "β", x)
   x <- gsub("gamma", "γ", x)
   #x <- gsub("delta", "δ", x)
   x <- gsub("epsilon", "ε", x)
   x <- gsub("zeta", "ζ", x)
   x <- gsub("ug/mL", "μg/mL", x) 
   x <- gsub("uL", "μL", x)
   x <- gsub("uM", "μM", x)
}

```

```{r import facs data & metadata, include = FALSE}

ExpDate <- "120901"

d.meta <- read_xlsx(file.path(getwd(), "example FACS data.xlsx"), sheet = "meta_data")
d.rawfacsdata <- read_xlsx(file.path(getwd(), "example FACS data.xlsx"), sheet = "raw_data")

```

```{r munging}

# Extract Diva Specimen information so that the flowjo data can be joined to the metadata
# remove extraneous stats from FlowJo output

names(d.rawfacsdata)[1] <- "V1"

not_all_na <- function(x) any(!is.na(x))

D.rn <- d.rawfacsdata %>%                                          
  mutate(ID_Tag = str_extract(V1, "(\\d+).fcs")) %>%
  mutate(ID_Tag = str_remove(ID_Tag, ".fcs")) %>%
  select_if(not_all_na) %>%
  mutate_at(vars(ID_Tag), as.numeric) %>%
  mutate(Specimen = map_chr(V1, ~ str_split(.x, "_")[[1]][2])) %>%    # Pull out DIVA "specimen" information
  mutate_at(vars(Specimen), as.numeric) %>%
  arrange(Specimen, ID_Tag) %>%
  filter(!grepl('FMO', V1)) %>%
  filter(V1 != "SD") %>%
  filter(V1 != "Mean") %>%
  mutate(ID_Tag = Specimen)


d.j <- left_join(d.meta, D.rn, by = "ID_Tag")
```

```{r reshape data}

# find the facs parameters using agrep on forward scatter (fsc) and side scatter (ssc)
facs_gates <- agrep(pattern = "fsc ssc", names(D.rn), max.distance = 0.1, value = TRUE, ignore.case = TRUE)

D.l <- d.j %>%                           #Convert to long format for downstream munging & plotting
  gather(key = "FACS_Param", value = "Readout", facs_gates)

```

```{r cleanup of FACS parameter names}

#  Change the FACS data labeling for final plotting
#  eg change "FSC SSC/Singlets/Live | Mean (Antigen A)"   # flowjo gate name
# into   "MFI Antigen A"                                  # a conventional FACS analyte label for the final plot

D.p <- D.l %>%
  mutate(param = str_extract(FACS_Param, "(?<=\\|)[^\\/]+")) %>%                  
  mutate(param = gsub(pattern = "Mean", replacement = "MFI", x = .$param)) %>%   
  mutate(param = gsub(pattern = "\\(", replacement = "", x = .$param)) %>%       
  mutate(param = gsub(pattern = "\\)", replacement = "", x = .$param)) %>%        
  mutate(Gate = str_extract(FACS_Param, "(?<=\\/)[^\\/]+(?=\\|)")) %>%
  mutate(ParentGate = str_extract_all(.$FACS_Param, regex("(?<=\\/)[^\\/]+(?=\\/)", multiline = TRUE))) %>%
  mutate(ParentGate = unlist(unique(map(.[,"ParentGate"], function(x) {map(x, last)})))) %>%
  mutate(param =  str_replace_all(.$param, "Freq. of Parent %", paste0(Gate, " (% freq. of ", ParentGate, ")")))

```

```{r more munging: remove white space & add greek letters}

#  Clean-up: trim white space & convert greek letters (eg uM to µM)
library(memisc)
library(dplyr)

D.p <- D.p %>% mutate(dplyr::across(where(is.character), trimws))                     # Remove white space
names(D.p) <- map_chr(names(D.p), ~ trimws(.x)) 

D.p <- D.p %>% mutate(across(where(is.character), toGreek))                                   # add Greek letters
names(D.p) <- as.list(names(D.p)) %>% map_chr(., ~ toGreekLiteral(.x))

write_xlsx(x = D.p, path = paste0(ExpDate, " FACSstats CleanedMerged.xlsx"), col_names = TRUE) # export table
```

```{r Plotting maturation status markers: Antigens A & B by tx group}

p.antigenA <- D.p %>% filter(param %in% "MFI Antigen A") %>%
  ggplot(., aes(`Experimental group`, Readout)) +
  geom_bar(stat = "summary", fun = "mean", alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2, color = "black", shape = 21, aes(fill = factor(`Donor`))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.subtitle = element_text(color = "ivory", hjust = 1, size = 8),
        plot.margin = unit(c(0.2, 0.2, 0.2 ,0.5), "in")) +                                       
  stat_summary(position = position_dodge(width = 0.9), fun.data=mean_se, fun.args = list(mult=1),
        geom="errorbar", color="black", width=0.2, size = 0.5) +
  stat_summary(position = position_dodge(width = 0.9), fun="mean", geom="point", color="black") +
  labs(fill = "Donor", 
       y = "MFI Antigen A", 
       x = NULL,
       title = "Stim B drives optimal \nupregulation of \nmaturity marker Antigen A") +
  scale_y_continuous(labels = label_number_si())

p.antigenB <- D.p %>% filter(param %in% "MFI Antigen B") %>%
  ggplot(., aes(`Experimental group`, Readout)) +
  geom_bar(stat = "summary", fun = "mean", alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2, color = "black", shape = 21, aes(fill = factor(`Donor`))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.subtitle = element_text(color = "ivory", hjust = 1, size = 8),
        plot.margin = unit(c(0.2, 0.2, 0.2 ,0.5), "in")) +                                       
  stat_summary(position = position_dodge(width = 0.9), fun.data=mean_se, fun.args = list(mult=1),
        geom="errorbar", color="black", width=0.2, size = 0.5) +
  stat_summary(position = position_dodge(width = 0.9), fun="mean", geom="point", color="black") +
  labs(fill = "Donor", 
       y = "MFI Antigen B", 
       x = NULL,
       title = "Stim C & D show the \nhighest expression of \nimmature cell marker Antigen B") +
  scale_y_continuous(labels = label_number_si())


```

```{r Plotting expression of each individual antigen by tx group, include = FALSE}

antigens <- D.p$param %>% unique

plots_to_export <- map(antigens, ~ D.p %>% filter(param %in% .x) %>%
  ggplot(., aes(`Experimental group`, Readout)) +
  geom_bar(stat = "summary", fun = "mean", alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2, color = "black", shape = 21, aes(fill = factor(`Donor`))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.subtitle = element_text(color = "ivory", hjust = 1, size = 8),
        plot.margin = unit(c(0.2, 0.2, 0.2 ,0.5), "in")) +                                             
  stat_summary(position = position_dodge(width = 0.9), fun.data=mean_se, fun.args = list(mult=1),
        geom="errorbar", color="black", width=0.2, size = 0.5) +
  stat_summary(position = position_dodge(width = 0.9), fun="mean", geom="point", color="black") +
  labs(fill = "Donor", 
       y = paste(.x), 
       x = NULL) +
#  scale_fill_viridis_d(option = "magma") +
  scale_y_continuous(labels = label_number_si())) %>% 
    marrangeGrob(., ncol = 2, nrow = 3)

ggsave("antigen responses to drug treatments.pdf", plots_to_export, height = 11, width = 8.5)

```

```{r Reshaping data & running PCA using tidymodels}

d.matrix <- D.p %>% 
  mutate(`Experimental group` = paste0(`Experimental group`, "_Donor ", Donor)) %>%
  dplyr::select(`Experimental group`, `param`, `Readout`) %>%
  pivot_wider(names_from = param, values_from = Readout) %>%
  unnest(-`Experimental group`)
  
pca_rec <- recipe(~., data = d.matrix) %>%         # recipe for pre-processing data...inclued data to be trained
  update_role(`Experimental group`, new_role = "id") %>%        # specify that it's the countries that should be clustered
  step_normalize(all_predictors()) %>%             # same as scale...mean = 0, SD = 1
  step_pca(all_predictors())                       # PCA for all predictors

pca_prep <- prep(pca_rec)                          # train a data recipe

```

```{r Plot of PCA results, include = FALSE}
##  
p.pca1_2 <- bake(pca_prep, new_data = NULL) %>% # apply a trained data recipe, don't change the data
  separate(., `Experimental group`, into = c("Treatment", "Donor"), sep = "_") %>%
  ggplot(aes(PC1, PC2, fill = `Treatment`)) +         
  geom_point(size = 2, shape = 21, color = "black",
             aes(color = `Treatment`, 
                 shape = `Treatment`), show.legend = FALSE)

p.pca3_4 <- bake(pca_prep, new_data = NULL) %>% # apply a trained data recipe, don't change the data
  separate(., `Experimental group`, into = c("Treatment", "Donor"), sep = "_") %>%
  ggplot(aes(PC3, PC4, fill = `Treatment`)) +         
  geom_point(size = 2, shape = 21, color = "black",
             aes(color = `Treatment`, 
                 shape = `Treatment`))

p.pca <- p.pca1_2 + p.pca3_4 & 
  geom_point(size = 2, shape = 21, color = "black", aes(color = `Treatment`, shape = `Treatment`), show.legend = FALSE) &
  plot_annotation(title = "Relationships among drug treatment groups of Cell Type X ", 
                  subtitle = "Dimensionality reduction (PCA) using Antigens A thru I",
                  caption = paste("n = 8 healthy donors per tx group; experiment from ", ExpDate))

```

```{r Plot loadings}

#  Which antigen contribute most to variation in each principal component

pca_comps <- tidy(pca_prep, 2) %>%            # tidy the output of a recipe
  filter(component %in% paste0("PC", 1:4)) %>%
  group_by(component) %>%      # select most important components of variation for each PC
  slice_max(abs(value), n = 7) %>%    
  ungroup()

p.loadings <- pca_comps %>%        
  mutate(value = abs(value),
         terms = str_remove(terms, "MFI ")) %>%
  group_by(terms) %>%
  ggplot(aes(value, fct_rev(terms), fill= `terms`)) +
  geom_col(position = "dodge", alpha = 0.7, color = "black", show.legend = FALSE) +
  facet_wrap(~ component, scales = "free_y") +
  labs(y = NULL, fill = NULL,
       x = "Absolute value contribution",
       title = "Top 7 antigens contributing to PC variation") +
  scale_fill_viridis_d()

```



#### Experimental Question:  How does treatment with drugs A-D affect differentiation of Cell Type X?

- In healthy human peripheral blood, antigens A & B indicate mature & immature forms of Cell Type X, respectively

- Problem: In disease, there is a defect in maturation of Cell Type X, and therapeutic intervention is needed to induce maturation of Cell Type X

- Goal: Assess the capacity of drug stimuli A, B, C, & D to induce maturation, as measured by FACS immunophenotyping

<br><br>


#### Goal 1: Assess which functional markers appear with drug-induced maturation of cell type X

- Antigen A indicates advanced maturation of Cell Type X

- Antigen B indicates immature cell status

<br><br>

#### Goal 2: Determine how drug combinations affect the essential cell functions performed by Antigens C - I

- Antigens C - I are functional markers associated with effector function of mature Cell Type X in healthy individuals. 

<br><br>

#### Experimental Setup
##### 1) Isolated immature cell type X from patients with disase and treated cells w/ drug stims A - D for 48 hrs @37degreesC
##### 2) Stained cells with antibodies against Antigens A - I
##### 3) Ran flow cytometry to visualize the phenotypic relationships of multiple cell type X treatment groups. 


<br><br>

#### Analysis Results

```{r, fig.align = 'center', fig.width = 8}
library(knitr)
p.pca
```

<br><br>

#### The PCA plot (PC1 x PC2) above suggests a possible differentiation path of Cell Type X.
This hypothesis can be confirmed by asking which treatment group expresses the highest level of Antigen A (maturation marker)?

```{r, fig.align = 'center', fig.width = 8}
p.antigenA + p.antigenB
```
##### Conclusions:
 - The principal components (PC1xPC2) dot plot above shows a differentiation path. The bar chart above shows that Stim C & D yields the most immature state of Cell Type X, & Stims B & C showing the greatest degree of maturation of Cell Type X

<br><br>

#### Does drug-induced maturation generate full effector function?
#### Which cell functional markers may be defective with drug-induced maturation?

- The differentiation path of Cell Type X is most clearly observable in PC1 & PC2. 
- Several antigens indicate fully developed effector function, including Antigens C, D, G, I, & J
- Which functional markers (Antigens C-I) are most strongly associated with drug-induced maturation? 

<br><br>

#### Plotting the PCA loadings to determine which functional antigens are most strongly associated with maturation path seen in PC1 & PC2.

```{r, fig.align = 'center', fig.width = 8}
p.loadings
```

#### Conclusions: Cell Type X treatment groups shown in the PC1 x PC2 plot show a developmental path strongly associated with changes in functional markers.
##### Among functional antigens, C,G, & J contribute most to PC1 and Antigens E & H contribute most to PC2. 
##### Thus, the maturation induced by stims B & C induces functional markers B,C,G,J,E,&H, but NOT functional marker F. 

<br><br>

##### Note:
 - Drug effects on functional markers can be confirmed in more detail by looking at supplementary exported data: 
 "antigen responses to drug treatments.pdf"
 - For example, Antigen J expression patterns are similar to previously defined maturation marker B. Antigen F does not track well with maturation. 

<br><br>

### In summary, the analysis finds that therapeutic treatment with drugs B & C can induce partial maturation of Cell Type X
- 48 hour treatment with drug stims B & C showed the strongest induction of classic markers of maturation (eg Antigen A). 
- Cell Type X treated with drugs B & C show some, but not all functional markers typically associated with maturation (Antigens C - I).  







